---
title: "Installing the crisprVerse and packages necessary for the tutorials"
author: Jean-Philippe Fortin, Luke Hoberecht
output: 
  github_document:
    toc: true
---

```{r, echo=FALSE, results="hide"}
options("knitr.graphics.auto_pdf"=TRUE)
```



# Installation 

We show in this tutorial how to install the crisprVerse packages, as well
as other packages
necessary for some of the [crisprVerse tutorials](https://github.com/crisprVerse/Tutorials).



## Requirements

The crisprVerse is supported for macOS, Linux and Windows machines.
It requires R version >=4.2.1. Some of the third-party functionalities are not
available for Windows machines (BWA alignment, and some of the scoring 
functions). To download and install R, see 
the [R-project website](https://www.r-project.org/).

## Bioconductor versions

The Bioconductor project has 2 concurrent versions: `release` and `devel`.
Currently (August 2022), the release version is`3.15`, and the
devel version is `3.16`. Release versions are created twice a year.

The current version of the crisprVerse was developed on the devel version
of Bioconductor (`3.16`) to make sure it accesses all of the latest 
developments. Earlier versions of some of our packages are available 
on the release version, but we do not recommend using the release version
as most of the functionalities described in the tutorials require
devel functionalities.

See the [Bioconductor install page](https://www.bioconductor.org/install/) for
more information re. Bioconductor. 


## Installing the core crisprVerse packages

The [crisprVerse package](https://github.com/crisprVerse/crisprVerse) installs
the core crisprVerse packages in a single command from Bioconductor. Simply
type in the following commands in an R session:


```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(version="devel")
BiocManager::install("crisprVerse")
```

Note that we specify the devel branch of Bioconductor so that we
can use the latest functionalities.


This will install the following packages:

* [crisprBase](https://github.com/crisprVerse/crisprBase) to specify and manipulate CRISPR nucleases.
* [crisprBowtie](https://github.com/crisprVerse/crisprBowtie) to perform gRNA spacer sequence
alignment with Bowtie.
* [crisprScore](https://github.com/crisprVerse/crisprScore) to annotate gRNAs with on-target
and off-target scores.
* [crisprDesign](https://github.com/crisprVerse/crisprDesign) to design and manipulate gRNAs
with `GuideSet` objects. 
* [crisprScoreData](https://github.com/crisprVerse/crisprScoreData) to use pre-trained models for the `crisprScore` package.

The following command will load all of those packages in an R session:

```{r, eval=FALSE}
library(crisprVerse)
```

You can check that all crisprVerse packages are up-to-date with 
`crisprVerse_update()`:

```{r, eval=FALSE}
crisprVerse_update()
```

## Installing data packages

The following genome data packages from Bioconductor are required 
for several of the tutorials:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(version="devel")
BiocManager::install("BSgenome.Mmusculus.UCSC.mm10")
BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")
BiocManager::install("BSgenome.Hsapiens.UCSC.hg38.dbSNP151.major")
BiocManager::install("BSgenome.Hsapiens.UCSC.hg38.dbSNP151.minor")
```

The [crisrpDesignData](https://github.com/crisprVerse/crisprDesignData) package 
is also required for most of the tutorials and can be installed directly 
from our GitHub page using the `devtools` package:


```{r, eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")

devtools::install.packages("crisprVerse/crisprDesignData")
```

## Installing optional packages

For maxOS and Linux users, the 
[crisprBwa](https://github.com/crisprVerse/crisprBwa) can be installed
from Bioconductor using the following:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(version="devel")
BiocManager::install("crisprBwa")
```





# Reproducibility

```{r}
sessionInfo()
```


# Reproducibility

```{r}
sessionInfo()
```

# References
