---
title: "Installing the crisprVerse and packages necessary for the tutorials"
author: Jean-Philippe Fortin, Luke Hoberecht
output: 
  github_document:
    toc: true
---

```{r, echo=FALSE, results="hide"}
options("knitr.graphics.auto_pdf"=TRUE)
```



# Installation 

We show in this tutorial how to install the crisprVerse packages, as well
as other packages
necessary for some of the [crisprVerse tutorials](https://github.com/crisprVerse/Tutorials).



## Requirements

The crisprVerse is supported for macOS, Linux and Windows machines.
It requires R version >=4.2.1. Some of the third-party functionalities are not
available for Windows machines (BWA alignment, and some of the scoring 
functions). To download and install R, follow the instructions on the  [R-project website](https://www.r-project.org/).

## Bioconductor versions

The Bioconductor project has 2 concurrent branches: `release` and `devel`.
Currently (August 2022), the release branch is `3.15`, and the
devel branch is `3.16`. Release versions are created twice a year. See the 
[Bioconductor install page](https://www.bioconductor.org/install/) for
more information regarding Bioconductor versions.

The crisprVerse ecosystem is currently available on the Bioconductor
devel branch (`3.16`). Earlier versions of some of our packages are available 
on the release branch, but we do not recommend using the release branch
as most of the functionalities described in the tutorials require updated
functionalities only available on the devel branch. 


## Installing the core crisprVerse packages

Type in the following commands in an R session to install the core
crisprVerse packages from the Bioconductor devel branch:


```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(version="devel")
BiocManager::install("crisprVerse")
```


This will install the following packages:

* [crisprBase](https://github.com/crisprVerse/crisprBase) to specify and manipulate CRISPR nucleases.
* [crisprBowtie](https://github.com/crisprVerse/crisprBowtie) to perform gRNA spacer sequence
alignment with Bowtie.
* [crisprScore](https://github.com/crisprVerse/crisprScore) to annotate gRNAs with on-target
and off-target scores.
* [crisprDesign](https://github.com/crisprVerse/crisprDesign) to design and manipulate gRNAs
with `GuideSet` objects. 
* [crisprScoreData](https://github.com/crisprVerse/crisprScoreData) to use pre-trained models for the `crisprScore` package.

The following command will load all of those packages in an R session:

```{r, eval=FALSE}
library(crisprVerse)
```

You can check that all crisprVerse packages are up-to-date with 
`crisprVerse_update()`:

```{r, eval=FALSE}
crisprVerse_update()
```

## Installing data packages

The following genome data packages from Bioconductor are required 
for several of the tutorials:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(version="devel")
BiocManager::install("BSgenome.Mmusculus.UCSC.mm10")
BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")
BiocManager::install("BSgenome.Hsapiens.UCSC.hg38.dbSNP151.major")
BiocManager::install("BSgenome.Hsapiens.UCSC.hg38.dbSNP151.minor")
```

The [crisrpDesignData](https://github.com/crisprVerse/crisprDesignData) package 
is also required for most of the tutorials and can be installed directly 
from our GitHub page using the `devtools` package:


```{r, eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")

devtools::install.packages("crisprVerse/crisprDesignData")
```

## Installing optional packages

For maxOS and Linux users, the 
[crisprBwa](https://github.com/crisprVerse/crisprBwa) can be installed
from Bioconductor using the following:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(version="devel")
BiocManager::install("crisprBwa")
```

The [crisprViz](https://github.com/crisprVerse/crisprViz) package is
currently under review at Bioconductor, but can be installed directly 
from GitHub:

```{r, eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")

devtools::install.packages("crisprVerse/crisprViz")
```



# Reproducibility

```{r}
sessionInfo()
```


# References
