---
title: "Building a gene annotation object"
output: 
  github_document:
    toc: true
# bibliography: references.bib
---

```{r, echo=FALSE, results="hide"}
options("knitr.graphics.auto_pdf"=TRUE)
```

Authors: Jean-Philippe Fortin, Luke Hoberecht

Date: `r format(Sys.time(), '%d %B, %Y')`


# Introduction

In this tutorial we describe the process for making and using a gene annotation object, which is required input for certain `crisprDesign` functions such as `queryTxObject` and `addGeneAnnotation` (it is also referred to as a `txObject`, after the argument name the object is passed to for applicable functions). This object is a `GRangesList` that is composed from a `TxDb` object to provide convenience and speed when retrieving gene annotation with `crisprDesign` functions. While a `TxDb` can be passed to `txObject` arguments, we recommend constructing and using a gene annotation object described herein. We will also describe the process for constructing and using a transcription start site (TSS) annotation object (similarly also referred to as a `tssObject`), which contains TSS coordinates and is a `GRanges` rather than a `GRangesList`.


# Installation

## Software requirements

### OS Requirements

This package is supported for macOS, Linux and Windows machines.
It was developed and tested on R version 4.2.

## Installation

This tutorial will use the `crisprDesign` package to demonstrate how to construct and retrieve information from gene annotaion objects. Examples using such objects will take advantages of the premade gene annotation objects in the `crisprDesignData` package. Both packages can be installed by typing the following commands in an R session:

```{r, warning=FALSE, message=FALSE, results='hide'}
install.packages("devtools")
devtools::install_github("Jfortin1/crisprDesign")
devtools::install_github("Jfortin1/crisprDesignData")
```


### Getting started

The packages can be loaded into an R session in the usual way:

```{r, warning=FALSE, message=FALSE, results='hide'}
library(crisprDesign)
library(crisprDesignData)
```

# Building gene annotation objects

The following demonstration will first create a `TxDb` object from the Ensembl database with `getTxDb` from `crisprDesign`, which uses the `GenomicFeatures` function `makeTxDbFromEnsembl`, before converting it to a gene annotation object. Note that this method requires an internet connection. If you have a General Feature Format (GFF) file from which you want to construct the gene annotation object, you can pass this to the `file` argument of the `crisprDesign` function `getTxDb`; this will create the `TxDb` object using the `GenomicFeatures` function `makeTxDbFromGFF`.


## Building a txObject

In this example, we construct a gene annotation object for the human genome, Ensembl release 104 (hg38). We then show how to subset this object to only contain annotation for our gene of interest, IQSEC3, although this method also works for an arbitrary list of genes. Subsetting is useful if memory constraints are a concern and only annotation for the gene(s) of interest is(are) desired.

```{r, warning=FALSE, message=FALSE, results='hide'}
library(GenomicRanges)

## genome-wide annotation
txdb <- getTxDb(organism="Homo sapiens",
                release=104)
grList <- TxDb2GRangesList(txdb)

## subset for IQSEC3
meta <- metadata(grList)
gene_id <- "ENSG00000120645" #IQSEC3
grListExample <- lapply(grList, function(gr){
    gr[gr$gene_id %in% gene_id]
})
grListExample <- GRangesList(grListExample)
metadata(grListExample) <- meta
GenomeInfoDb::genome(grListExample) <- "hg38"
```

Let's look at our gene annotation object:

```{r}
names(grListExample)
grListExample$transcripts
```

## Building a tssObject

Building a `tssObject` requires only one additional step after constructing the `txObject` described above, with the convenience function `getTssObjectFromTxObject` in `crisprDesign`. Let's apply the function to our subset example:

```{r}
tssObjectExample <- getTssObjectFromTxObject(grListExample)
tssObjectExample
```


# Using gene annotation objects

Gene (or TSS) annotation objects are necessary for the full characterization of guide-RNAs (gRNAs) and serve as input for several `crisprDesign` functions, including `queryTxObject`, `queryTssObject`, `addGeneAnnotation`, `addTssAnnotation`, and `addSpacerAlignments`. The following sections demonstrate the information it contains and how to use the object. We will take advantage of precomputed gene annotation objects available in the `crisprDesignData` package:

|  Object name | Object class | Version | Description |
|----------- | ----------- | ----------- |----------- | 
| `txdb_human`  | `GRangesList`   | Release 104 | Ensembl gene model for human (hg38/GRCh38) |
| `txdb_mouse`  | `GRangesList`   | Release 102 | Ensembl gene model for mouse (mm10/GRCm38) |
| `tss_human`   | `GRanges`       | Release 104 | Ensembl-based TSS coordinates for human (hg38/GRCh38) |
| `tss_mouse`   | `GRanges`       | Release 102 | Ensembl-based TSS coordinates for human (mm10/GRCm38) |


## Using txObjects

Let's first look at the `txdb_human` object.

```{r, warning=FALSE, message=FALSE}
data(txdb_human, package="crisprDesignData")
```

We can look at metadata information about the gene model by using the `metadata` function from the `S4Vectors` package:

```{r, warning=FALSE, message=FALSE, collapse=TRUE}
head(S4Vectors::metadata(txdb_human))
```

The object is a `GRangesList` with 7 elements that contain genomic coordinates for different levels of the gene model:

```{r, collapse=TRUE}
names(txdb_human)
```

As an example, let's look at the `GRanges` containing genomic coordinates for all exons represented in the gene model:

```{r, collapse=TRUE}
txdb_human$exons
```

The function `queryTxObject` in `crisprDesign` is a user-friendly function to work with such objects. For instance we can find the CDS coordinates for the KRAS transcripts using the following lines of code:

```{r, warning=FALSE, message=FALSE, collapse=TRUE}
cds <- queryTxObject(txdb_human,
                     featureType="cds",
                     queryColumn="gene_symbol",
                     queryValue="KRAS")
head(cds)
```

We can also pass `txdb_human` to the `txObject` parameter of `addGeneAnnotation` (required) and `addSpacerAlignments` (optional) to add gene annotation information to a `GuideSet`, provided the `genome` of the `GuideSet` is also `hg38`. Below is a simple example using `txdb_human` in `addGeneAnnotation`:

```{r, warning=FALSE, message=FALSE, results='hide'}
library(BSgenome.Hsapiens.UCSC.hg38)
gs <- findSpacers(head(cds),
                  crisprNuclease=SpCas9,
                  bsgenome=BSgenome.Hsapiens.UCSC.hg38)
gs <- head(gs)
```

```{r, collapse=TRUE}
## add gene annnotation
gs1 <- addGeneAnnotation(gs,
                         txObject=txdb_human)
geneAnnotation(gs1)
```

For examples using a gene annotation object in `addSpacerAlignments`, see the tutorials for [CRISPRko design with Cas9](https://github.com/crisprVerse/Tutorials/tree/master/Design_CRISPRko_Cas9) or [CRISPRko design with Cas12a](https://github.com/crisprVerse/Tutorials/tree/master/Design_CRISPRko_Cas12a).


## Using tssObjects

The `tss_human` and `tss_mouse` objects are `GRanges` representing the TSS coordinates for human and mouse, respectively. The coordinates were extracted from the transcripts stored in the Ensembl-based models `txdb_human` and `txdb_mouse` using the function `getTssObjectFromTxObject` from `crisprDesign`.

Let's take a look at `tss_human`:

```{r, collapse=TRUE}
data(tss_human, package="crisprDesignData")
head(tss_human)
```


The function `queryTss` in `crisprDesign` is a user-friendly function to work with such objects, accepting an argument called `tss_window` to specify a number of nucleotides upstream and downstream of the TSS. This is particularly useful to return genomic regions to target for CRISPRa and CRISPRi. 

For instance, if we want to target the region 500 nucleotides upstream of any of the KRAS TSSs, one can use the following lines of code:

```{r, warning=FALSE, message=FALSE, collapse=TRUE}
library(crisprDesign)
tss <- queryTss(tss_human,
                queryColumn="gene_symbol",
                queryValue="KRAS",
                tss_window=c(-500,0))
tss
```

Similar to `txObject`s and the `addGeneAnnotation` function, `tssObject`s can be passed to the `crisprDesign` function `addTssAnnotation`, and is also an optional argument for `addSpacerAlignments`. The functions append TSS annotation to the `GuideSet` with respect to gRNAs and their alignments, respectively; see the tutorials [CRISPRa design](https://github.com/crisprVerse/Tutorials/tree/master/Design_CRISPRa) or [CRISPRi design](https://github.com/crisprVerse/Tutorials/tree/master/Design_CRISPRi) for more information.


# License

The package is licensed under the MIT license. 

# Reproducibility

```{r}
sessionInfo()
```


