---
title: "gRNA design for CRISPR interference (CRISPRi)"
author: Jean-Philippe Fortin, Luke Hoberecht
output: 
  github_document:
    toc: true
bibliography: references.bib
---

```{r, echo=FALSE, results="hide"}
options("knitr.graphics.auto_pdf"=TRUE)
```



# Introduction

This tutorial will demonstrate how to use `crisprDesign` to design gRNAs 
for CRISPR interference (CRISPRi). Specifically, we will target the 
human KRAS gene and use the SpCas9 nuclease.

# Installation

See the [Installation tutorial](https://github.com/crisprVerse/Tutorials/tree/master/Installation) to learn how to install the packages necessary for this tutorial:
`crisprDesign`, `crisprDesignData`


# Terminology

See the [CRISPRko design tutorial](https://github.com/crisprVerse/Tutorials/tree/master/Design_CRISPRko_Cas9) to get familiar 
with the terminology used throughout this tutorial.


# CRISPRi design

For CRISPR activation (CRISPRa) and interference (CRISPRi) applications,
the CRISPR nuclease is engineered to lose its endonuclease activity, 
and should therefore not introduce double-stranded breaks (DSBs). 
We will use the dead SpCas9 (dSpCas9) nuclease as an example here. 
Note that users don't have to distinguish between dSpCas9 and SpCas9 
when specifying the nuclease in the crisprVerse as they 
do not differ in terms of the characteristics stored in 
the `CrisprNuclease` object.

In CRISPRi, fusing dSpCas9 with a Krüppel-associated box (KRAB) domain has been
shown to be effective at repressing transcription in mammalian cells
[@crispri]. The dSpCas9-KRAB fused protein is a commonly-used construct to
conduct CRISPR inhibition (CRISPRi) experiments. To achieve optimal inhibition,
gRNAs are usually designed targeting the region directly downstream of the gene
transcription starting site (TSS).

`crisprDesign` provides functionalities to be able to take into account 
design rules that are specific to CRISPRi applications. The `queryTss` function
allows for specifying genomic coordinates of promoter regions. 
The `addTssAnnotation` function annotates gRNAs for known TSSs, 
and includes a column `dist_to_tss` that gives the distance in
nucleotides between the TSS position and the PAM site of the gRNA. 
For CRISPRi, we recommend targeting the region 25-75bp region 
downstream of the TSS for optimal inhibition; see @sanson2018optimized for 
more information. Finally, the function `addCrispraiScores` adds 
CRISPRi-specific on-target activity scores based on the work of [@crisprai]. 


## Creating the GuideSet

We first start by loading the required packages:

```{r, message=FALSE, warning=FALSE, results='hide' }
library(crisprBase)
library(crisprDesign)
library(crisprDesignData)
library(BSgenome.Hsapiens.UCSC.hg38)
```


To demonstrate CRISPRi design, we will design gRNAs to inhibit
expression of the human KRAS 
gene using the SpCas9 nuclease. To accomplish this, we want our gRNAs to
target the region downstream of the KRAS TSS; let's consider the window
containing 500bp immediately downstream of the TSS to explore candidate gRNAs.

We first need to retrieve the TSS coordinates for KRAS. 
These data are conveniently stored in the 
`crisprDesignData` package as the dataset `tss_human`. For more information 
on `tss_human` and how to create similar TSS annotation objects, 
see the [Building a gene annotation object](https://github.com/crisprVerse/Tutorials/tree/master/Building_Gene_Annotation) 
tutorial.


We load the TSS coordinates stored in the `tss_human` object

```{r, warning=FALSE, message=FALSE, results='hide'}
data("tss_human", package="crisprDesignData")
```

and query for KRAS using the `queryTss` function from `crisprDesign`:

```{r, warning=FALSE, message=FALSE, results='hide'}
target_window <- c(0, 500)
target_region <- queryTss(tss_human,
                          queryColumn="gene_symbol",
                          queryValue="KRAS",
                          tss_window=target_window)
```

```{r, collapse=TRUE}
target_region
```

We load the `crisprNuclease` object storing information about the
SpCas9 nuclease from the `crisprBase` package:

```{r}
data(SpCas9, package="crisprBase")
```

We then find all candidate protospacer sequences in our target region 
with `findSpacers`:

```{r, warning=FALSE, message=FALSE, results='hide'}
gs <- findSpacers(target_region,
                  crisprNuclease=SpCas9,
                  bsgenome=BSgenome.Hsapiens.UCSC.hg38)
```

```{r, collapse=TRUE}
gs
```


## Annotating the GuideSet

Next, we annotate our candidate gRNAs to assess quality. There are several 
functions in `crisprDesign` that provide annotation for features that are 
not specific to CRISPRi, for which we refer the reader to the
[CRISPRko design with Cas9](https://github.com/crisprVerse/Tutorials/tree/master/Design_CRISPRko_Cas9) tutorial for more information. The sections below will 
cover annotation functions that are of particular interest to 
CRISPRi applications.


## Adding TSS annotation

As the name implies, the `addTssAnnotation` function annotates gRNAs with 
TSS context such as the distance between the gRNA and the TSS, as well as 
which TSS is targeted (many genes contain different TSSs corresponding to
different isoforms). 

The function requires a `tssObject` object, and the `tss_window` values 
that we used earlier to define the target region. We can then retrieve 
the appended annotation with the accessor function `tssAnnotation`:

```{r, collapse=TRUE}
gs <- addTssAnnotation(gs,
                       tssObject=tss_human,
                       tss_window=target_window)
tssAnnotation(gs)
```


## Adding spacer alignments with TSS annotation

As with all CRISPR applications, potential off-targets effects are an 
important concern in assessing gRNA quality. While this concern is 
somewhat moderated for CRISPRi, since the dead CRISPR nuclease does 
not make DSBs, we should be aware of off-targetsoccuring in the promoter regions of other genes. This can be handled by  passing our `tssObject` to the `addSpacerAlignments` function. We will search  for up to 2 mismatches and increase the size of our `tss_window` (which defines the promoter region when searching for off-targets) to err 
on the safe side. 

Similar to the CRISPRko design tutorial, we need to specify a Bowtie index
of the human referenge genome; see the [Building genome indices for short read aligners](https://github.com/crisprVerse/Tutorials/tree/master/Building_Genome_Indices) tutorial to learn how to create such an index. 

Here, we specify the index that was available to us when generating this
tutorial:

```{r}
# Users need to specify the path of their bowtie index
index_path <- "/Users/fortinj2/crisprIndices/bowtie/hg38/hg38"
```

We are ready to add on- and off-target alignments:

```{r, warning=FALSE, message=FALSE, results='hide'}
gs <- addSpacerAlignments(gs,
                          aligner="bowtie",
                          aligner_index=index_path,
                          bsgenome=BSgenome.Hsapiens.UCSC.hg38,
                          n_mismatches=2,
                          tssObject=tss_human,
                          tss_window=c(-500, 2000))
```
```{r, collapse=TRUE}
gs
```

Including a `tssObject` parameter in the `addSpacerAlignments` function 
appends columns to the `GuideSet` that tallies the alignments restricted 
to the defined (via `tss_window`) promoter regions: 
`n0_p`, `n1_p`, and `n2_p` (the `_p` suffix denotes "promoter").


## Adding CRISPRai scores

The CRISPRai algorithm was developed by the Weissman lab to score SpCas9 
gRNAs for CRISPRa and CRISPRi applications for the 
human genome [@crisprai]. The function `addCrispraiScores` 
implements this algorithm to add scores to the `GuideSet`.
Compared to other on-target scoring algorithms, it requires several additional
inputs:

- The `gr` argument is the `GRanges` object derived from the `queryTss`
function and used to create the `GuideSet` object. 
In our example, this is the object named `target_region`.
- The `tssObject` argument is a `GRanges` object that contains 
TSS coordinates and annotation. It must also contain the following 
columns: `ID`, `promoter`, `tx_id`, and `gene_symbol`.
Our `tssObject` in this instance is `tss_human`.
- `geneCol` indicates which column of `tssObject` should be used as the 
unique gene identifier.
- `modality` is the modality of the CRISPR application, in our case, `CRISPRi`.
- `fastaFile` is the path of a FASTA file containing the sequence of the
human reference genome in hg38 coordinates. This file is available [here](https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.fa.gz).
- `chromatinFiles` is a vector of length 3 specifying the path of files 
containing the chromatin accessibility data needed for the algorithm 
in hg38 coordinates. The chromatin files can be downloaded from Zenodo [here](https://zenodo.org/record/6716721#.YrzCfS-cY4d).


We first prepare all needed inputs for `addCrispraiScores`.
We start by specifying the location of the FASTA file on our local machine:

```{r}
fastaPath <- "/Users/fortinj2/crisprIndices/genomes/hg38/hg38.fa"
```

This corresponds to the path where the downloaded file from [here](https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.fa.gz) is
stored. Next, we specify the location of the chromatin files:

```{r}
mnasePath <- "/Users/fortinj2/crisprIndices/chromatin/hg38/crispria_mnase_human_K562_hg38.bigWig"
dnasePath <- "/Users/fortinj2/crisprIndices/chromatin/hg38/crispria_dnase_human_K562_hg38.bigWig"
fairePath <- "/Users/fortinj2/crisprIndices/chromatin/hg38/crispria_faire_human_K562_hg38.bigWig"
chromatinFiles <- c(mnase=mnasePath,
                    dnase=dnasePath,
                    faire=fairePath)
```

This should correspond to the files that were downloaded from 
[here](https://zenodo.org/record/6716721#.YrzCfS-cY4d).

We are now ready to add the scores:

```{r, warning=FALSE, message=FALSE, results='hide', eval=TRUE}
results <- addCrispraiScores(gs,
                             gr=target_region,
                             tssObject=tss_human,
                             geneCol="gene_id",
                             modality="CRISPRi",
                             fastaFile=fastaPath,
                             chromatinFiles=chromatinFiles)
```

Let's look at the results:

```{r, collapse=TRUE, eval=TRUE}
results
```

You can see that the column `score_crispri` was added to the `GuideSet`.
Note that this function works identically for CRISPRa applications, with 
the `modality` argument replaced by `CRISPRa`. 



# Session Info

```{r}
sessionInfo()
```


# References
